Effects of experiencing the COVID-19 pandemic on optimistically biased belief updating

Version 1

Abstract

Abstract

Optimistically biased belief updating is essential for mental health and resilience in adversity. Here, we asked how experiencing the COVID-19 pandemic affected optimism biases in updating beliefs about the future. One hundred and twenty-three participants estimated the risks of experiencing adverse future life events in the face of beliefdisconfirming evidence either outside the pandemic (n=58) or during the pandemic (n=65). While belief updating was optimistically biased and Reinforcement-learning-like outside the pandemic, the bias faded, and belief updating became more rational Bayesian-like during the pandemic. This malleability of anticipating the future during the COVID-19 pandemic was further underpinned by a lower integration of positive belief-disconfirming information, fewer but stronger negative estimations, and more confidence in base rates. The findings offer a window into the putative cognitive mechanisms of belief updating during the COVID-19 pandemic, driven more by quantifying the uncertainty of the future than by the motivational salience of optimistic outlooks.

Introduction

Anticipating the future is an essential part of human thinking. These beliefs guide how we understand the world; updating them is vital for learning to make better predictions and to generalize across contexts. Interestingly, belief updating tends to be optimistically biased(1). Even when confronted with negative information, humans often downplay its importance to maintain an optimistic view of the future. The underlying mechanism of optimistically biased belief updating involves an asymmetry in learning from positive and negative belief-disconfirming information(2)(3)(4). This optimism bias can be formalized in two ways(5). Reinforcement learning (RL) models assume that belief updating is proportional to the estimation error, which reflects a form of surprise when we face information that disconfirms our initial beliefs. For example, the positive surprise of learning that the risk of contracting a disease is lower than initially thought implies a motivational salience, guiding toward lowering the initial risk estimation and avoiding worsening beliefs(5).

On the contrary, many aspects of human reasoning, including belief updating, can be formalized using Bayes’ rule. Bayesian updating models assume that initial beliefs are prior probabilities or guesses that express how strongly a person predicts future life events within a finite set of possible outcomes before being presented evidence in line and at odds with this initial belief. Following Bayes’ rule, a belief is updated by combining the observed evidence for the initial belief with its likelihood ratio, which expresses the evidence’s diagnosticity(6). For example, updating one’s risk for contracting a disease in a Bayesian way means quantifying the uncertainty of this event for oneself compared to others. This can be optimal and non-biased, which we refer to here as rational Bayesian. On the contrary, it can be optimistically biased, with more Bayesian beliefs updating when the odds against contracting a disease are favorable.

Although RL-like and Bayesian updating models make different assumptions about the updating strategy, they are complementary and powerful formalizations of human reasoning. Both models provide insight into hidden, latent variables of the updating process. Most notably, the learning rate indicates how quickly a person updates their beliefs. It comprises two free parameters: The scaling parameter weighs how much the estimation error is considered following an RL-like updating model or the product of the evidence for the prior and its diagnosticity following a Bayesian updating model. The asymmetry parameter expresses how much this scaling differs in the face of positive and negative belief-disconfirming evidence. Both the scaling and asymmetry parameters can vary between individuals and contexts and, through this variance, offer possible explanations for the idiosyncrasy in belief-updating behavior and its deviation from rationality.

The COVID-19 pandemic represented a chronic, adverse life context, drastically altering individual and social realities. The existing literature documenting the impact of the COVID-19 pandemic and the associated changes in everyday life on mental health has shown that stress, anxiety, and depression have increased during the pandemic(7)(8). Moreover, previous work has demonstrated that belief updating during reversal learning became more erratic and was linked to a rise in paranoia during the COVID-19 pandemic across the US(9). However, it is unknown how optimism biases in belief updating about the future and their underlying putative mechanisms changed during the experience of such an unprecedented, adverse life event. Our hypothesis was twofold. We argued that maintaining optimistically biased belief updating under lasting, adverse life conditions is adaptive. The optimism bias benefits exploratory behavior, reduces stress, and improves mental and physical health and well-being(10)(11)(12)(13)(14)(15). These benefits promote resilience, especially for fitness and survival during a pandemic(15)(16) (17).

On the contrary, optimism biases can lead to suboptimal decision-making. Contextual factors such as acute stress, perceived threat, and depression have been shown to reduce or even reverse optimistically biased belief updating(17)(18)(19)(20)(21)(22)(23). These findings suggest that optimistically biased belief updating should be weaker when experiencing a pandemic.

We leveraged a belief-updating dataset from 123 participants tested between 2019 and 2022 to rule between these alternative hypotheses. Among them, fifty-eight participants were tested outside the context of the COVID-19 pandemic, either in October 2019, three months before the outbreak in France (n=30) or two years later in June 2022 (n=28), after the lift of the sanitary state of emergency. Their belief updating behavior was compared to sixty-five participants tested during the sanitary state of emergency due to the COVID-19 outbreak in France. This was either during the first very strict lockdown of social and economic life (e.g., schools closed, stay-at-home orders, shops, and museums closed) from March to April 2020 (n=34) or one year later in Mai 2021 (n=31), when lockdowns were less strict (e.g., schools open, museums and shops closed, part-time curfew), but the COVID-19 pandemic still unfolding. Belief updating was measured by a behavioral task that asked participants to estimate their risk of experiencing adverse future life events before and after receiving information about these events’ actual base rates. Observed belief updating behavior was fitted to an RL-like and a Bayesian updating model to gain insight into potential underlying strategies of belief updating. The learning rates were compared across groups for insight into how experiencing the COVID-19 pandemic changed beliefs about the future and their updating in the face of belief-disconfirming evidence.

Methods

Ethical considerations

The study protocol followed the Declaration of Helsinki and was approved by the local ethics committee at Sorbonne University. All participants provided informed consent. The authors declare no competing interests.

Participants

One hundred twenty-five participants (mean age = 37.50 ± 1.28, 99 females) allotted to four different groups were recruited for the study (see Table 1 and SI Table 15) via a public advertisement. Two participants from the group tested in June 2022 were excluded from the analyses because they always indicated the same risk estimate for each event.

Experimental design

As shown in Figure 1a, a first group of 30 participants (mean age = 33.73 ± 1.96, 18 females) was recruited in October 2019 before the COVID-19 outbreak in France. These participants were tested in the laboratory. A second group of 34 participants (mean age = 42.24 ± 3.34, 25 females) was recruited for online testing during the first COVID-19-related lockdown of social and economic life, with schools closed from March to April 2020. A third group of 31 participants (mean age = 42.42 ± 3.35, 20 females) was recruited and tested online immediately after the last lockdown and still during the COVID-19 pandemic in Mai 2021. A fourth group of 30 participants (mean age = 34.66 ± 2.71, 16 females) was recruited at the lift of the COVID-19 pandemic-related state of emergency and tested in the laboratory in June 2022. This group was also used to rule out an eventual effect of task design. Half of them performed a one-run (n = 15) task design, and the other half performed a two-run (n = 15) task design (e.g., see in more detail the belief updating task description below). Note the 30 participants tested before the COVID-19 pandemic were recontacted during the first strict lockdown to re-perform the belief updating task online. This allowed us to check for the effects of experiencing a COVID-19-related lockdown within the same cohort of participants. Two of the thirty participants of this group did not respond. Therefore, the sample size for the within-group test-retest analyses was 28 participants.

Experimental design.

(a) Timeline of testing. Four groups were tested before the COVID-19 outbreak in October 2019, during the first complete lockdown of social and economic life in March and April 2020, after a partial lockdown in Mai 2021, and after the pandemic in June 2022. (b) Belief updating task. Panels show subsequent events within a good news trial (left panels) and a bad news trial (right panel). Responses were self-paced. The task goal was to estimate the risk of experiencing different adverse future life events (e.g., tooth decay) for oneself and for somebody else before (E1) and after (E2) being presented with information about the event’s prevalence in the general population (i.e., base rate).

Sample sizes:

The sample sizes were determined by a power analysis using the power curve function in R (version 1.2.5033) and building on the good news/bad news bias observed in the first group tested in October 2019 before the COVID-19 outbreak in France. The sample size required to replicate a significant effect of estimation error valence on the absolute updating with a power between 80% and 90% lay between 28 and 35 participants, respectively.

Belief Updating Task

All participants performed a belief-updating task (Figure 1b). For in-person testing, stimulus presentation and response recording were done with the Psychophysics toolbox in MATLAB (R2018b, Update 6, version 9.5.0.1265761). The online testing was done using Qualtrics (Qualtrics Software, version March 2020 of Qualtrics, Copyright 2020 Qualtrics. Available at https://www.qualtrics.com).

The task comprised 40 trials with 40 adverse lifetime events and base rates. In each trial, participants were asked to estimate the likelihood of experiencing an adverse event in the future for themselves and somebody else before and after receiving information about the likelihood of occurrence in the general population (i.e., the base rate). The adverse life events and their actual base rates were taken from previously published work in healthy controls(2). The base rates were uniformly distributed and ranged between 10 and 70%. Participants were instructed to rate their estimates between 3% and 77%.

The task design variations:

Fifty-eight participants underwent assessment outside the COVID-19 pandemic, with 45 performing a two-run task design (n=30 tested before the outbreak in October 2019; n = 15 tested at the end of the sanitary state of emergency in June 2022). The remaining 13 participants tested outside the pandemic performed the one-run task design like the 65 participants tested during the pandemic.

In the two-run task design, participants performed a first run of 40 trials. Each trial started with the display of an adverse lifetime event. Participants were asked to estimate the risk of experiencing this event in the future for themselves (E1 rating) and for somebody else (eBR rating). At the end of each trial, they received information about the event’s base rates and rated their confidence. In a second run, they saw an adverse future life event and its base rate on each trial. They then re-estimated their risk (E2 rating) on a trial-bytrial basis. The one-run task design is displayed in Figure 4b and consists of one run of 40 trials. Within each trial, participants first estimated the risk of experiencing a future adverse lifetime event for themselves (E1 rating) and for somebody else (eBR rating), were presented with the base rate for this event (BR), rated their confidence in the base rate and re-estimated their risk of experiencing the event in the future (E2 rating). Note that all analyses were controlled for these differences in task design, which had non-significant effects on belief updating, confidence ratings, estimation error magnitude, and learning rates (see SI tables 2, 3, 5, 6, 7).

Belief updating task measures of interest:

The estimation error indicated whether participants overestimated or underestimated their likelihood of experiencing an adverse event (E1) relative to its actual base rate (aBR). The estimation error (EE) was calculated according to the equation i: EE=E1−aBR i.

The estimation error was further used to categorize trials into good or bad news trials:

For good news trials, the estimation error was positive (EE > 0), which indicated an overestimation of one’s likelihood of experiencing an adverse life event relative to the base rate of that event (E1 > aBR). For bad news trials, the estimation error was negative (EE < 0), which indicated an underestimation of one’s likelihood of experiencing an adverse event relative to its actual base rate (E1 < aBR).

The main variable of interest was the magnitude of belief updating (UPD), which was calculated as the difference between the first (E1) and the second (E2) estimate after receiving information about the base rate (BR). Notably, the update was calculated for good and bad news trials, respectively, following equation ii: f EE > 0 UPDgood news=E1−E2if EE < 0 UPDbad news=E2−E1 ii.

Lastly, the difference between updating after good and bad news was calculated to assess the updating bias following equation iii: UDB=UPDgood news−UPDbad news iii.

A positive difference indicated that participants updated their beliefs about their lifetime risks of experiencing adverse life events more frequently following good news than bad news.

For each participant, trials that did not receive a response (on average of 0.44 trials per subject) and trials with an EE = 0 (on average of 0.63 trials per subject) were excluded from the analyses.

The distance measured the extent to which participants consider their probability of experiencing a given adverse event (E1) different from the lifetime risk of someone from a similar socio-economic background (eBR). If positive, it reflected an optimistic bias in initial estimates. The following, distance = eBR — E1, was calculated.

Model-free statistical analyses of observed belief updating behavior:

The main aim of this study was to assess how belief updating was affected by the context of experiencing the COVID-19 pandemic.

We, therefore, conducted between-context analyses, contrasting groups tested during (i.e., during the first lockdown in March/April 2020 and immediately after the last lockdown one year later in Mai 2021) and outside the pandemic context (i.e., before the outbreak in October 2019 and one year after the pandemic in June 2022). All statistical tests were conducted using the MATLAB Statistical Toolbox (MATLAB 2018b, MathWorks) and JASP (JASP 0.16.4).

A first linear mixed effects model (LME 1) was fitted to the absolute belief updating, following equation iv: | UPD |=β0 intercept+βEE| EE |+βEE valence EE valence + βcontextcontext + βdesigndesign +βageage +βgendergender +βeducationeducation +βEE valence*contextEE valence by context +(1|subject)+(1+|EE| |subject)+(1+EE valence | subject) iv.

The model included fixed effects for estimation error magnitude (|EE|), estimation error valence (EEvalence, coded -1 for bad news trials and 1 for good news trials), context (coded 0 for outside and 1 for during the COVID-19 pandemic), task design (coded 1 for one-run, 2 for two-run design), age, gender (coded 0 for male, 1 for female participants), level of education, and the interaction of interest EE valence by context. The model also included random intercepts nested by subject number and random slopes for estimation error magnitude and valence.

Subsequently, the same Linear Mixed Effects (LME) model was applied again to the absolute belief update to explore the categorical effect of context in conjunction with EE valence. This allowed for a more specific comparison of the impact of EE valence between contexts (groups): those tested before the COVID-19 pandemic outbreak in October 2019 (baseline) compared to those tested during the initial COVID-19-related lockdown (context 1), those tested immediately after the last lockdown during the pandemic in Mai 2021 (context 2), and those tested one year post-pandemic in June 2022 (context 3), respectively.

Post-hoc two-tailed and one-tailed t-tests were conducted to characterize the directionality of detected main effects and interactions.

Model-based analyses of belief-updating behavior

To gain more insight into putative cognitive mechanisms of belief updating during and outside the COVID-19 pandemic, two families of non-linear computational models were fitted to observe belief updating behavior, which is specified below.

Model specifications

(1) Reinforcement learning model of belief updating.

A Reinforcement learning-like model assumed that belief updating is proportional to the magnitude of the estimation error following Kuzmanovic and Rigoux, 2017 (5). The learning rate scaled the effect of the estimation error on belief updating following the generic equation v: UPD = LR * EE v.

Importantly, the learning rate was estimated for good and bad news trials separately and following equations vi and vii: LRgood news=S + A vi. LRbad news=S − A vii.

For both types of trials, the learning rate was composed of two components that varied across participants. The scaling parameter (S) measured the extent to which a participant took the prediction error into account when updating beliefs. The asymmetry parameter (A) indicated to what extent the belief updating differed for positive and negative estimation errors. The priors for scaling and asymmetry were untransformed and unbound. The mean of the prior distribution for scaling was set to 1. Thus, a scaling of 1 meant that the updating magnitude equaled the estimation error magnitude. The mean of the prior distribution for the asymmetry parameter was set to zero. An asymmetry parameter value larger than zero meant positively biased updating, whereas an asymmetry parameter smaller than zero meant negatively biased belief updating.

A version of the RL-like model of belief updating took the personal relevance (PR) of presented adverse future life events into account following equations viii and ix: UPDgood news=(alpha+asymmetry)* EE * (1 - PR) viii. UPDbad news=(alpha−asymmetry)* EE * (1 - PR) ix.

The PR weighed the estimation error (EE) and corresponded to the difference between the estimated base rate (eBR — the estimated risk for somebody else) and the initial estimate (E1 — the estimated risk for oneself). Based on the sign of this difference between eBR and E1, the PR was calculated following equations x to xii: if eBR <E1 PR=eBR−E1eBR−1 x. if eBR <E1 PR=E1−eBR99−eBR xi. if eBR=E1 PR=0 xii.

(2) Bayesian belief updating model

A second family of computational models was fitted to belief updating behavior and assumed that belief updating was proportional or equal to the Bayes rule, following equations xiii and xiv: UPDgood news=(E1−E2b)*(scaling+asymmetry) xiii. UPDbad news=(E2b−E1)*(scaling+asymmetry) xiv.

The scaling parameter (S) corresponded to the tendency of participants to update their beliefs in response to the presented base rate. A scaling smaller than one (S < 1) indicated lesser belief updating than what was predicted by the Bayes rule, and a scaling larger than one (S > 1) indicated more updating than predicted by the Bayes rule.

The Bayes rule was used to define a Bayesian second estimate (E2b, the updated belief), which was calculated following equations xv and xvi: E2b=Prior * LHR xv. E2b=Prior ∗ LHR1+(Prior ∗ LHR) xvi.

With the Prior = P(BR), corresponding to the base rate (BR) of each event following equation xvii: Prior =BR1−BR xvii.

The Likelihood Ratio (LHR) indicates the probability of the initial estimate (E1) relative to the likelihood of the alternative estimated base rate (eBR) following equation xviii: LHR=E11−E1eBR1−eBR xviii.

Alternative models of these two model families (RL and Bayesian) were fitted to the observed belief-updating behavior. Each model alternative represented a different combination of free parameters composing the learning rate to test a total of 12 assumptions about the cognitive process underlying belief updating:

Belief updating is asymmetrical and proportional to the estimation error: S + A varied across participants.

Belief updating is non-asymmetrical and proportional to the estimation error: S varied, A was silent (fixed to zero)

Asymmetrical updating is equal to the estimation error: S is fixed (to 1), and A is varied

Updating equals the estimation error: S and A were fixed

Belief updating is asymmetrical, proportional to the estimation error, and moderated by the personal relevance of events (PR): S + A vary. PR is weighting the EE following equations v. and vi.

Belief updating is non-asymmetrical and proportional to the estimation error moderated by PR: S varies, A is fixed, and PR weighs the EE following equations v. and vi.

Asymmetrical updating equal to the estimation error moderated by PR: S is fixed, A varies, and PR weighs the EE following equations v. and vi.

Updating equals the estimation error moderated by PR: S and A are fixed, and PR weighs the EE following equations v. and vi.

Belief updating is asymmetrical and proportional to Bayes rule: S and A vary.

Belief updating is proportional to a rational Bayes rule: S varies, and A is fixed.

Belief updating equals an asymmetrical Bayes rule: S is fixed, and A varies.

Belief updating equals a rational Bayes rule: S and A are fixed.

Model estimation

Models were estimated following the procedure reported by Kuzmanovic and Rigoux 2017 and Bottemanne et al., 2022(5)(21). In short, models were not hierarchical, and parameter estimation was thus less sensitive to differences in group sample sizes. For each participant, optimal scaling and asymmetry parameter values were obtained using Bayesian variational inferences implemented in the VBA toolbox(24).

Model comparisons

The free energy approximations for a model’s evidence in each participant were entered into a random effect Bayesian model comparison that yielded the two criteria considered for model selection: the estimated model frequency (Ef) in each group and the exceedance probability (pxp), which corresponded to the likelihood of the model to describe best the observed belief updating behavior in each participant.

Parameter recovery

Parameter recovery analysis was conducted to check whether the free parameters of the winning models were identifiable and described the data better than any other set of parameters. The procedure was the same as reported in Bottemanne et al. 2022(21). In short, to validate the accuracy of the fitting procedure in providing meaningful parameter values, simulated belief updating data was generated using the observed parameter values for both the optimistic RL model and the optimistic Bayesian updating model. Subsequently, we applied the fitting procedure to these simulated data to iteratively ‘recover’ the parameters. Thereby, the means of the parameters were set to correspond to the observed sample means (i.e., scaling = 0.37 ± 0.02, asymmetry = 0.06 ± 0.01 for the RL model; scaling = 0.40 ± 0.02, asymmetry = 0.04 ± 0.01 for the Bayesian model). This process was iterated to simulate 40 values of belief updates 123 times. The model was then inverted by fitting it to the simulated data, yielding a new set of recovered values for scaling and asymmetry. Finally, the recovered and estimated parameters were compared by assessing their correlation using Pearson’s correlation coefficients.

Parameter comparisons

To compare learning rates and learning rate components across groups, we used the parameters from the optimistically biased RL-like model (RL model 1), which performed best when fitted to the whole dataset (Ef = 0.40, pxp = 1).

Individual learning rates from this RL model 1 and their scaling and asymmetry components were the dependent variables (DV) of the following generic linear mixed effects model (equation xix and xx): LR=β0 Intercept + βvalence valence + βcontext context + βdesign design + βage age + βgender gender +βeducation education +βvalence*context valence by context +(1|subject) xix. LR components = β0 Intercept + βcontext context + βage age + βgender gender + βeducation education + (1|subject) xx.

The model included fixed effects for news valence (valence, coded 1 for good news, -1 for bad news), context (coded 0 for outside the pandemic, 1 for during the pandemic), task design (coded 1 for one-run and 2 for two-run), age, gender (coded 0 for male, 1 for female participants), and level of education. It also tested the interaction of interest context by valence. The intercept was nested at random by subject number.

Post-hoc one-sampled and two-sampled t-tests were conducted to characterize the directionality of effects.

Results

Effects of experiencing the COVID-19 pandemic on the optimism bias in belief updating

A linear mixed effects (LME) model was fitted to absolute belief updates to test whether belief updating was less or more biased during the COVID-19 pandemic. The model found a significant interaction estimation error valence by context (β = -5.54, SE = 1.69, t(232) = -3.28, p = 0.001, 95% CI [-8.87 – -2.21]; SI table 2). As shown in Figures 1a and b, the optimism bias in belief updating disappeared during the COVID-19 pandemic compared to participants tested outside the pandemic. More specifically, it was decreased among participants tested during the initial COVID-19-related strict lockdown in March and April 2020 (EEvalence by context 1: β = -7.39, SE = 2.29, t(228) = -3.32, p = 0.002, 95% CI [-11.91 – -2.86]; SI table 3), as well as in Mai 2021 (EEvalence by context 2: β = -5.59, SE = 2.36, t(228) = -2.37, p = 0.02, 95% CI [-10.24 – -0.93]; SI table 3), compared to those tested before the outbreak in October 2019, respectively. The bias re-emerged among participants tested one year later at the time of the lift of the sanitary state of emergency in June 2022, returning to levels akin to those observed before the pandemic in October 2019 (EEvalence by context 3: β = -2.11, SE = 2.46, t(228) = -0.86, p = 0.39, 95% CI [-6.95 – 2.73]; Figure 2a, SI table 3). The effect of the COVID-19 pandemic on belief updating was driven by a significant decrease in belief updating following good news during the pandemic compared to participants tested outside the pandemic (t(121) = 2.66, p = 0.009, Cohen’s d = 0.48, two sampled, two-tailed t-test, Figure 2b). No contextual group difference was observed for belief updating following bad news (t(121) = -1.77, p = 0.08, Cohen’s d = -0.32, two sampled, two-tailed t-test, Figure 1b). This effect could be reproduced when fitting an analogous LME to belief updates observed in a group of participants (n=28) who were tested both before and during the pandemic (EE valence by context interaction: β = -7.66, SE = 1.49, t(103) = -5.13, p < 0.001, 95% CI [-10.62 – 4.70]; SI figure 1, SI table 4). Note that all effects were controlled for participant age, years of higher education, gender, confidence in the base rates, belief updating task design, and estimation error magnitude.

Behavioral results.

(a) Boxplots display the belief-updating bias (i.e., the difference between the belief update for good news minus belief update for bad news) in each of the four participant groups tested before the pandemic in October 2019 (n=30), during the first lockdown from March to April 2020 (n=34), with less restrictive measures in Mai 2021 (n=31), and at the end of the pandemic in June 2022 (n=28). (b) Belief updating for good and bad news during (n=65) and outside the pandemic (n=58). (c) Confidence ratings, and (d) estimation errors for bad and good news during and outside the pandemic. Boxplots in all panels display 95% confidence intervals, with boxes indicating the interquartile range from Q1 25th to Q3 75th percentile. The horizontal black lines indicate medians and whiskers range from minimum to maximum values and span 1.5 times the interquartile range. The dots correspond to individual participants. The source data file provides exact p-values. *p < 0.05 two-sampled, two-tailed t-tests, * p < 0.05 two-sampled, one-tailed t-tests.

Effects of COVID-19-related lockdown on belief updating variables

As shown in Figure 2c, experiencing the COVID-19 pandemic influenced participants’ confidence in the base rates, with significantly lower confidence ratings observed among those tested outside the pandemic compared to those tested during it (β = 14.11, SE = 4.52, t(233) = 3.12, p = 0.002, 95% CI [5.19 – 23.02]; SI Table 5). Moreover, a significant interaction of EE valence by context (β = 2.19, SE = 0.67, t(233) = 3.28, p = 0.001, 95% CI [0.88 – 3.51]; SI Table 6) was found for absolute estimation error magnitude. This finding indicated that participants tested during the pandemic overestimated their risk of experiencing adverse future life events relative to base rates more largely than participants tested outside the pandemic (t(121) = -3.01, p = 0.003, Cohen’s d = -0.54, Figure 2d). On the contrary, the two groups did not differ significantly in the magnitude of negative estimation errors (i.e., initial underestimations relative to base rates, t(121) = -0.49, p = 0.63, Cohen’s d = -0.09, two-sampled, two-tailed t-tests (Figure 2d). This finding contrasts with the observed difference in how often they made positive estimation errors (i.e., the number of good news trials). Participants tested during the pandemic overestimated less frequently than participants tested outside the pandemic (t(121) = 2.40, p = 0.02, Cohen’s d = 0.43). No significant difference between groups was found for the frequency of underestimations (i.e., reflected by the number of bad news trials; t(121) = -1.85, p = 0.07, Cohen’s d = -0.33). These results indicated that participants held fewer but more negative future outlooks during the pandemic compared to those tested outside the pandemic.

Effects of COVID-19-related lockdown on putative mechanisms of belief updating.

We then sought to identify which putative strategy participants used to update their beliefs about the future during and outside the pandemic. To answer this question, we used computational modeling and model comparisons to rule between twelve alternative models. This approach revealed that belief updating outside the pandemic was more RL-like and optimistic (pxp = 1, Ef = 0.77), while during the pandemic, it was best explained by a rational Bayesian updating model (pxp = 0.90, Ef = 0.43; Figure 3a-c). Similar findings were obtained when conducting model comparisons in the participants tested before and during the lockdown (n=28, SI Figure 2).

Computational model comparisons.

Twelve alternative models from RL-like (blue) and Bayesian (orange) updating model families were fitted to observed belief updates for participants tested during the COVID-19 pandemic (left panel columns) and outside the pandemic (right column panels). (a) Protected exceedance probabilities for each of the 12 alternative models. (b) Posterior model attributions. Colored cells display the probability that individual participants (y-axis) will be best explained by a model version (x-axis). (c) Estimated model frequencies correspond to how many participants are expected to be best described by a model version, with error bars corresponding to standard deviations. The red line indicates the null hypothesis that all model versions are equally likely in the cohort (chance level). Labels on the x-axis of the histogram and bar graphs indicate the model versions with non-silenced parameters (S— scaling, A— Asymmetry) and PR — personal relevance of events.

Effects of the Covid-19-related lockdown on hidden, latent variables of belief updating

Next, we compared the effects of experiencing the COVID-19 pandemic on the learning rates and its components.

A linear mixed effects model (LME), analogous to the LME fitted to observed belief updates, was fitted to the learning rates and detected a main effect of EE valence (β = 0.09, SE = 0.01, t(236) = 7.14, p = 1.18e-11, 95% CI [0.06 – 0.11]), and a significant interaction EE valence by context (β = -0.03, SE = 0.02, t(236) = -2.11, p = 0.04, 95% CI [-0.07 – -0.002]; Figure 4a, SI table 7). A main effect of EE valence (β = 0.07, SE = 0.02, t(105) = 3.22, p = 0.002, 95% CI [0.03 – 0.12]; SI table 8) and context (β = -0.10, SE = 0.03, t(105) = -3.10, p = 0.002, 95% CI [-0.17 – -0.04]; SI table 8) on learning rates was detected when comparing the participants, who were both tested before and during the pandemic. As shown in Figure 4a, all participants’ learning rates were lower in response to bad news than to good news. Still, the difference between good and bad news learning rates was significantly reduced for participants tested during the pandemic. In line with the observed belief updating after good and bad news, the effect of context on the learning rates was driven by a decrease in the learning rates from positive estimation errors in participants tested during the pandemic compared to participants tested outside the pandemic (t(121) = 2.17, p = 0.03, Cohen’s d = 0.39). Both groups did not differ in their learning rates from negative estimation errors (t(121) = 0.87, p = 0.39, Cohen’s d = 0.16, two sampled, two-tailed t-tests).

Parameter comparisons between participants tested during (n=65) and outside (n=58) the COVID-19 pandemic.

(a) Learning rates. Boxplots display 95% confidence intervals for learning rates from the RL-like updating model that assumed updating is proportional to the estimation error with an asymmetry and a scaling learning rate component. (b) Parameter recovery for learning rate components. Pearson’s correlation between generating and recovered parameters for scaling (left panel) and asymmetry (right panel) learning rate component. r—Pearson’s correlation coefficient against zero. Source data and exact p-values are provided as a Source Data file. (c) Group comparisons scaling and asymmetry components. Boxplots display 95% confidence intervals for the learning rate’s scaling (left panel) and the asymmetry (right panel) component. Boxes in all boxplots correspond to the interquartile range from Q1 25th percentile to Q3 75th percentile. The horizontal black lines indicate medians and whiskers range from minimum to maximum values and span 1.5 times the interquartile range. The dots correspond to individual participants. *p <0.05. P-values were obtained with two-sampled, two-tailed t-tests between groups, and exact p-values are provided in the source data file.

Parameter recovery was successful for the scaling and asymmetry components of the learning rates (Figure 4b), which indicated that the model gave identifiable values for these parameters. We, therefore, were able to explore potential group differences in the learning rate components in more detail. Linear mixed effects modeling found a main effect of context for the asymmetry component (β = -0.04, SE = 0.02, t(117) = -2.32, p = 0.02, 95% CI [-0.07 – -0.01]; Figure 4c, SI table 9), but not for the scaling component (β = -0.07, SE = 0.05, t(117) = -1.54, p = 0.13, 95% CI [-0.16 – 0.02]; Figure 3c, SI table 10). The average asymmetry of learning rates was positive in both groups but significantly smaller in participants tested during the pandemic than those tested outside (t(121) = 2.00, p =0.048, Cohen’s d = 0.36, two-sampled, two-tailed t-test, Figure 4c). This result indicated that participants considered positive estimation errors more than negative ones but less when experiencing the COVID-19 pandemic. Similar results were found in the within-subject group (n = 28), with a significant main effect of context on asymmetry (β = -0.06, SE = 0.02, t(51) = -3.72, p = 0.001, 95% CI [-0.09 – -0.03]; SI table 11), but not on alpha (β = -0.10, SE = 0.05, t(51) = -1.96, p = 0.06, 95% CI [-0.21 – 0.003]; SI table 12).

Discussion

This study investigated how experiencing the COVID-19 pandemic impacted the optimism bias in updating beliefs about the future. Belief updating was optimistically biased before the COVID-19 outbreak. It faded during the COVID-19 pandemic and reemerged after the pandemic. The lack of optimistically biased belief updating during the pandemic was related to three effects: (1) a decreased sensitivity to positive belief-disconfirming information, (2) fewer but stronger negative beliefs about the future, and (3) more confidence in base rates. Computational modeling showed that a rational Bayesian model best-described belief updating during the pandemic.

In contrast, an optimistic RL-like model best approximated belief updating outside the pandemic. Both models showed that the attenuated optimism bias in belief updating during the pandemic was not due to a learning deficit. The groups were similar in how much they integrated overall evidence in favor or against initial beliefs. On the contrary, it was explained by a diminished learning asymmetry in considering positive belief-disconfirming evidence that paralleled the observed belief-updating behavior.

The finding that optimistically biased belief updating faded during the pandemic favors the hypothesis that experiencing an adverse life event such as a pandemic weakens optimistic outlooks. It further aligns with the body of research that has explored the malleability of the optimistically biased belief updating and information integration under acute threat and stress and mood disorders such as depression(17)(19)(18)(20). While our findings align with these previous findings, we also observed a difference. Notably, our sample tested during the pandemic considered positive, favorable information less while showing no change in negative, unfavorable information consideration. Differences in populations and task designs might explain these odds. However, it could also be specific to experiencing the COVID-19 pandemic, which involved an immediate, unpredictable, and global health threat with high uncertainty about its outcome and significant psychological repercussions (7)(25).

Mental health assessments during the COVID-19 pandemic indicated that anxiety, stress, paranoia, and depression levels were more prevalent in the population(8)(9). The rapid spread of SARS-CoV-2 and the emergence of COVID-19 cases worldwide constitute a challenging situation that, in five months, has shifted from an elusive and distant threat to an immediate and drastic health and economic crisis. All citizens were confronted daily with alarming figures such as infection rates or mortality, and rather mundane everyday activities, from grocery shopping to jogging, became stressful and threatening situations during which one could catch a potentially fatal infection. Moreover, for many, the COVID-19-related lockdowns of social and economic life implied a physical cut-off from friends and relatives, and life plans, routines, and activities were severely disrupted. It also implied a substantial financial risk. Previous research has shown that economic uncertainties, particularly during marked economic inequality and epidemics, can contribute to beliefupdating fallacies reflected by the rise of conspiracy theories(26). We did not collect information about the COVID-19 infection status of participants, which precludes a direct exploration of the immediate effects of experiencing the infection on belief-updating behavior and the potential interaction with anxiety and stress levels. This limitation is noteworthy, as the impact of experiencing the pandemic on belief updating about the future could differ between those who directly experienced infection and those who remained uninfected. It is also important to acknowledge that our study was timely and geographically limited to the context of the COVID-19 outbreak in France. Cultural variations and differences in governmental responses to contain the spread of SARS-CoV-2 may have impacted the optimism biases differently.

The observed lack of optimistically biased belief updating may be interpreted as an adaptive response to the experience of an unprecedented level of uncertainty and chronic threat during a global crisis. Although we did not have access to anxiety and stress perceptions during and outside the pandemic, our computational modeling results corroborate to some extent this interpretation. Notably, belief updating was more optimistically biased RL-like outside the pandemic and more rational Bayesian-like during the pandemic. The biased RL-like updating behavior observed outside the pandemic indicated that participants relied on the motivational salience of positive estimation errors to teach them how to update their beliefs about the future by trial and error. This finding aligns with past work, showing that RL-like updating models best explain belief updating in non-threatening, non-stressful, and predictable laboratory contexts(5). It further suggests that RL strategies are a computationally efficient way to guide decision-making and belief formation when the environment is stable and predictable(27). For instance, in environments with well-defined reward structures, the human brain has been shown to efficiently rely on RL and avoid a computational overhead associated with the Bayesianlike inference process(28).

On the contrary, belief updating was more rational and Bayesian-like during the COVID-19 pandemic, indicating that participants weighed the uncertainty of evidence in favor of and against their prior beliefs. This finding aligns with research about Bayesian networks to model semantic knowledge processing under uncertainty ￼and with work that uses Bayes rules to understand how humans learn and choose under uncertainty(30)(31) (32).

It is essential to acknowledge that computational modeling provides insight into potential mechanisms, but this excludes inferences on whether humans indeed update beliefs in the way the best-fitting model assumes. Other models, such as evidence accumulation models, may also work when humans update their beliefs about the future and their immediate surroundings(33). Unfortunately, we did not assess reaction times during belief updating, crucial for fitting evidence accumulation models such as drift-diffusion models to observed behavior. However, we can infer from our findings that the two model families employed to fit observed belief-updating behavior represented two different but complementary prediction strategies. These strategies were then used to function in the uncertainty of real-life conditions. We call for more studies investigating these computational models’ psychological and biological validity under certainty and uncertainty.

In conclusion, our results provide insight into the resilience and adaptability of beliefupdating processes during and following the COVID-19 pandemic. They demonstrate the malleability of the human ability to anticipate the future and how it can adapt to real-life conditions under which an overly optimistic view of future risks would be harmful.

eLife Assessment

This important study addresses the question of how large-scale events such as the COVID-19 pandemic can change peoples' beliefs and their updates. Using a well-validated task, the authors find that belief updating becomes less optimistically biased during COVID-19, compared to prior to it. While they have solid evidence for specific changes of positive learning rates in their RL model and should be commended for more extensive modeling of the task than many previous studies, comprehensive model confusion and recovery analyses could further strengthen their claims by clarifying the specificity of their findings. As with many manipulations outside the experimenters' control, it remains unclear which psychological factor impacted by the pandemic drives the group differences, and sample sizes are by necessity on the smaller side as data cannot readily be acquired.

Reviewer #1 (Public review):

Summary:

This manuscript uses a well-validated behavioral estimation task to investigate the degree to which optimistic belief updating was attenuated during the 2020 global pandemic. Online participants recruited during and outside of the pandemic estimated how likely different negative life events were to happen to them in the future and were given statistics about these events happening. Belief updating (measured as the degree to which estimations changed after viewing the statistics) was less optimistically biased during the pandemic (compared to outside of it). This resulted from reduced updating from "good news" (better than expected information). Computational models were used to try to unpack how statistics were integrated and used to revise beliefs. Two families of models were compared - an RL set of models where "estimation errors" (analogous to prediction errors in classic RL models) predict belief change and a Bayesian set of models where an implied likelihood ratio was calculated (derived from participants estimations of their own risk and estimation of the base rate risk) and used to predict belief change. The authors found evidence that the former set of models accounted for updating better outside of the pandemic, but the latter accounted for updating during the pandemic. In addition, the RL model provides evidence that learning was asymmetrically positively biased outside of the pandemic but symmetric during it (as a result of reduced learning rates from good news estimation errors).

Strengths:

Understanding whether biases in learning are fixed modes of information processing or flexible and adapt in response to environmental shocks (like a global pandemic or economic recession) is an important area of research relevant to a wide range of fields, including cognitive psychology, behavioral economics, and computational psychiatry. The study uses a well-validated task, and the authors conduct a power analysis to show that the sample sizes are appropriate. Furthermore, the authors test that their results hold in both a between-group analysis (the focus of the main paper) and a within-group analysis (mainly in the supplemental).

The finding that optimistic biases are reduced in response to acute stress, perceived threat, and depression has been shown before using this task both in the lab (social stress manipulation), in the real world (firefighters on duty), and clinical groups (patients with depression). However, the work does extend these findings here in important ways:

(1) Examining the effect of a new real-world adverse event (the pandemic).

(2) The reduction in optimistic updating here arises due to reduced updating from positive information (previously, in the case of environmental threat, this reduction mainly arose from increased sensitivity to negative information).

(3) Leveraging new RL-inspired computational approaches, demonstrating that the bias - and its attenuation - can be captured using trial-by-trial computational modeling with separate learning rates for positive and negative estimation errors.

Weaknesses:

Some interpretation and analysis (the computational modeling in particular) could be improved.

On the interpretation side, while the pandemic was an adverse experience and stressful for many people (including myself), the absence of any measures of stress/threat levels limits the conclusions one can draw. Past work that has used this task to examine belief updating in response to adverse environmental events took physiological (e.g., SCR, cortisol) and/or self-report (questionnaires) measures of mood. In SI Table 1, the authors possibly had some questionnaire measures along these lines, but this might be for the participants tested during the pandemic.

On the analysis side, it was unclear what the motivation was for the different sets of models tested. Both families of models test asymmetric vs symmetric learning (which is the main question here) and have similar parameters (scaling and asymmetry parameters) to quantify these different aspects of the learning process. Conceptually, the different behavioral patterns one could expect from the two families of models needed to be clarified. Do the "winning" models produce the main behavioral patterns in Figure 1, and are they in some way uniquely able to do so, for instance? How would updating look different for an optimistic RL learner versus an optimistic Bayesian RL learner? Would the asymmetry parameter in the former be correlated with the asymmetry parameter in the latter? Moreover, crucially, would one be able to reliably distinguish the models from one another under the model estimation and selection criteria that the authors have used here (presenting robust model recovery could help to show this)?

Reviewer #2 (Public review):

The authors investigated how experiencing the COVID-19 pandemic affected optimism bias in updating beliefs about the future. They ran a between-subjects design testing for participants on cognitive tasks before, during, and after lifting the sanitary state of emergence during the pandemic. The authors show that optimism bias varied depending on the context in which it was tested. Namely, it disappeared during COVID-19 and re-emerged at the time of lift of sanitary emergency measures. Through advanced computational modeling, they are able to thoroughly characterize the nature of such alternations, pinpointing specific mechanisms underlying the lack of optimistic bias during the pandemic.

Strengths pertain to the comprehensive assessment of the results via computational modeling and from a theoretical point of view to the notion that environmental factors can affect cognition. However, the relatively small sample size for each group is a limitation. A major impediment interpreting of the findings is the need for additional measures. While the information on for example, risk perception or the need for social interaction was collected from participants during the pandemic, the fact that these could not be included in the analysis hinders the interpretation of findings, which is now generally based on data collected during the pandemic, for example, reporting increased stress. While authors suggest an interpretation in terms of uncertainty of real-life conditions it is currently difficult to know if that factor drove the effect. Many concurrent elements might have accounted for the findings. This limits understanding of the underlying mechanisms related to changes in optimism bias

Author response:

The following is the authors’ response to the original reviews

Reviewer #1:

Summary:

This manuscript uses a well-validated behavioral estimation task to investigate how optimistic belief updating was attenuated during the 2020 global pandemic. Online participants recruited during and outside of the pandemic estimated how likely different negative life events were to happen to them in the future and were given statistics about these events happening. Belief updating (measured as the degree to which estimations changed after viewing the statistics) was less optimistically biased during the pandemic (compared to outside of it). This resulted from reduced updating from "good news" (better than expected information). Computational models were used to try to unpack how statistics were integrated and used to revise beliefs. Two families of models were compared - an RL set of models where "estimation errors" (analogous to prediction errors in classic RL models) predict belief change and a Bayesian set of models where an implied likelihood ratio was calculated (derived from participants estimations of their own risk and estimation of the base rate risk) and used to predict belief change. The authors found evidence that the former set of models accounted for updating better outside of the pandemic, but the latter accounted for updating during the pandemic. In addition, the RL model provides evidence that learning was asymmetrically positively biased outside of the pandemic but symmetric during it (as a result of reduced learning rates from good news estimation errors).

Strengths:

Understanding whether biases in learning are fixed modes of information processing or flexible and adapt in response to environmental shocks (like a global pandemic or economic recession) is an important area of research relevant to a wide range of fields, including cognitive psychology, behavioral economics, and computational psychiatry. The study uses a well-validated task, and the authors conduct a power analysis to show that the sample sizes are appropriate. Furthermore, the authors test that their results hold in both a between-group analysis (the focus of the main paper) and a within-group analysis (mainly in the supplemental).

The finding that optimistic biases are reduced in response to acute stress, perceived threat, and depression has been shown before using this task both in the lab (social stress manipulation), in the real world (firefighters on duty), and clinical groups (patients with depression). However, the work does extend these findings here in important ways:

(1) Examining the effect of a new real-world adverse event (the pandemic).

(2) The reduction in optimistic updating here arises due to reduced updating from positive information (previously, in the case of environmental threat, this reduction mainly arose from increased sensitivity to negative information).

(3) Leveraging new RL-inspired computational approaches, demonstrating that the bias - and its attenuation - can be captured using trial-by-trial computational modeling with separate learning rates for positive and negative estimation errors.

Weaknesses:

Some interpretation and analysis (the computational modeling in particular) could be improved.

On the interpretation side, while the pandemic was an adverse experience and stressful for many people (including myself), the absence of any measures of stress/threat levels limits the conclusions one can draw. Past work that has used this task to examine belief updating in response to adverse environmental events took physiological (e.g., SCR, cortisol) and/or self-report (questionnaires) measures of mood. In SI Table 1, the authors possibly had some questionnaire measures along these lines, but this might be for the participants tested during the pandemic.

Thank you for this review.

We agree that the lack of physiological and self-report measures of stress, threat, and perceived uncertainty limits the interpretation of findings regarding potential psychological factors. Some self-reported anxiety and perceived risk measures experienced during the lockdowns were collected in a subset of participants (n=40, counting n=21 tested before and during the 1st strict lockdown, and n=19 tested solely during the 1st lockdown). These reports were given retrospectively at the time of release of the 1st lockdown in summer 2020 when the pandemic was still unfolding (SI Table 1).

Exploratory correlations revealed some noteworthy trends. We found that participants who reported to have perceived a bigger risk of death due to contagion were also those who were less optimistically biased when updating their beliefs about adverse future life risks during the first strict COVID-19-related lockdown (Pearson’s r = -0.36, p = 0.02).

Moreover, parameter estimates from the computational models of belief updating showed associations with specific survey responses: The rational Bayesian model’s scaling parameter correlated positively with adherence to distancing measures (r = 0.41, p = 0.01) and negatively with the need for social contact (r = -0.37, p = 0.02). This result indicated that participants who were updating their beliefs faster were more likely to follow preventive guidelines and felt less social craving. Meanwhile, the asymmetry parameter correlated negatively with mask wearing (r = -0.41, p = 0.01), positively with physical contact with close others (r = 0.32, p = 0.04) and satisfaction with social interactions (r = 0.33, p = 0.04). This suggests that participants who displayed some asymmetry in belief updating during the COVID-19 pandemic were less likely to comply with mask-wearing rules and more likely to engage in social interactions.

However, these results did not survive correction for multiple comparisons and the sample size for correlational analyses is in the lower range. The subjective measures of anxiety and fear of contagion did not significantly correlate to the updating bias, or any other variable measured by the belief updating task (e.g. estimation error, updating magnitude).

We now further discuss on page 12 the limitation, which reads:

“We did not collect physiological measures of stress or information about the COVID-19 infection status of participants, which precludes a direct exploration of the immediate effects of experiencing the infection on belief-updating behavior and the potential interaction with anxiety and stress levels. Although subjective ratings of the perceived risk of death from COVID-19 correlated negatively to the beliefs updating bias measured during the pandemic, this result was obtained retrospectively in a subset of participants (SI section 4). We thus cannot directly attribute the observed lack of optimistically biased belief updating during the lockdown to psychological causes such as heightened anxiety and stress. This limitation is noteworthy, as the impact of experiencing the pandemic on belief updating about the future could differ between those who directly experienced infection and those who remained uninfected. It is also important to acknowledge that our study was timely and geographically limited to the context of the COVID-19 outbreak in France. Cultural variations and differences in governmental responses to contain the spread of SARS-CoV-2 may have impacted the optimism biases in belief updating differently.”

On the analysis side, it was unclear what the motivation was for the different sets of models tested. Both families of models test asymmetric vs symmetric learning (which is the main question here) and have similar parameters (scaling and asymmetry parameters) to quantify these different aspects of the learning process. Conceptually, the different behavioral patterns one could expect from the two families of models needed to be clarified.

Thank you for raising this point. We agree that a clearer conceptual distinction between the two model families can help strengthen the interpretation of our findings. We have added the following considerations to the introduction on pages 2–3, which now reads:

“The underlying mechanism of optimistically biased belief updating involves an asymmetry in learning from positive and negative belief-disconfirming information[2,3,4], which can unfold in two ways following Reinforcement learning (RL) or Bayes rule[5].

Conceptually, Reinforcement learning (RL) and Bayesian models of belief updating are complementary but make different assumptions about the hidden process humans may use to adjust their beliefs when faced with information that contradicts them. The RL models assume belief updating is proportional to the estimation error. The key idea of the estimation error expresses the difference between how much someone believes they will experience a future life event and the actual prevalence of the event in the general population. This difference can be positive or negative. A scaling and an asymmetry parameter quantify the propensity to consider the estimation error magnitude and its valence, respectively. These two free parameters form the learning rate, which indicates how fast and biased participants update their beliefs.

In contrast, Bayesian models assume that following Bayes’ rule the posterior, updated belief is a new hypothesis, formed by pondering prior knowledge with new evidence. The prior knowledge consists in information about the prevalence of life events in the general population. The new evidence comprises various alternative hypotheses. It examines how likely a specific event is to occur or not occur for oneself, compared to the likelihood that it will happen or not happen to others. This probabilistic adjustment of beliefs about future life events can be considered as an approximation of a participant’s confidence in the future. The two free parameters of the Bayesian belief updating model scale how much the initial belief deviates from the updated, posterior belief (i.e., scaling parameter) and the propensity to consider the valence of this deviance (i.e., asymmetry parameter).

Although RL-like and Bayesian updating models make different assumptions about the updating strategy, they are complementary and powerful formalizations of human reasoning. Both models provide insight into hidden, latent variables of the updating process. Most notably, the learning rate and its components, the scaling and asymmetry parameters, which can vary between individuals and contexts and, through this variance, offer possible explanations for the idiosyncrasy in belief-updating behavior and its cognitive biases. “

Do the "winning" models produce the main behavioral patterns in Figure 1, and are they in some way uniquely able to do so, for instance? How would updating look different for an optimistic RL learner versus an optimistic Bayesian RL learner?

We now show that the winning models can reproduce the main behavioral patterns (revised Figure 1b).

Moreover, we plotted estimated and observed average belief updating for each participant (n=123) using the overall best-fitting asymmetrical RL-like updating model shown in SI Figure 6.

Would the asymmetry parameter in the former be correlated with the asymmetry parameter in the latter? Moreover, crucially, would one be able to reliably distinguish the models from one another under the model estimation and selection criteria that the authors have used here (presenting robust model recovery could help to show this)?

The asymmetry parameter estimated with the optimistically biased RL- and Bayesian models did correlate (r = 0.735; p < 0.001).

However, we argue that while the observed updating behavior and estimated free parameters are similar for RL-like and Bayesian learners, the underlying assumed cognitive processes differed and are identifiable. To test this assumption, we have added a model recovery analysis now reported in the supplement section 2c and main manuscript’s methods section pages 24–25.

As shown in SI Figure 5 confusion matrix, there is evidence for strong recovery of nearly all models, and importantly for the two winning models: the optimistically biased RL-like model and the rational Bayesian model of belief updating. This analysis thus rules out that the two model families were confused and mitigate concerns about the validity of the model selection.

Note, one exception was observed. The RL-like and Bayesian updating models that assumed no scaling and asymmetry were best recovered by their respective models that estimated the asymmetry parameter. Many factors could explain this. For example, it could be that the models, which assumed asymmetry, but no scaling, may have captured some bias in updating due to noise generated by the zero parameter models.

A justification is also needed to focus on the "RL-like updating model with an asymmetry and scaling learning rate component" in Figure 3. As I understand it, this model fits best outside of the pandemic, but another model - the Rational Bayesian Model - does worse (and does the best during the pandemic). What model best combines the groups (outside and inside the pandemic)?

We thank the reviewer for highlighting the need to justify our focus on the biased RL-like updating model in Figure 3. The model chosen for parameter comparison was selected based on a model comparison procedure conducted across all 12 models, including data from all participants (both those tested outside and during the pandemic, n=123). This model comparison revealed that Model 1 — the RL model with both asymmetry and scaling learning rate parameters estimated — provided the best fit across the entire dataset (Ef = 0.40, pxp = 0.99). As such, we focused on this model for parameter comparisons in Figure 3 to ensure consistency with the model comparison results and to interpret the parameters in the context of the overall best-fitting model. We added this information on top of the model parameter comparison results on page 8. Moreover, SI Figure 6 in the supplements shows how this model reproduces the observed belief updating in each of the 123 participants.

Why do the authors use absolute belief updating (|UPD|) in the first linear mixed effects model (equation iv)? Since an update is calculated differently depending on whether information calls for an update in an upward or downward direction, I do not understand the need to do this (and it means that updates that go in the wrong direction - away from the information - are counted as positive)

Thank you for driving our attention to this point. The ‘absolute belief updating’ note was incorrect, and we apologize for the confusion. To be precise, we did not use absolute updating values in our analyses. Belief updating was assumed on each trial to go either toward the base rate (e.g., Update = E2 – E1) for negative estimation errors or away from it for positive estimation errors (e.g., Update = E1 – E2). Updates that went in the wrong direction, further away from the base rate, were thus counted and included in the analysis with their negative sign. We have corrected this important point in equation iv of the methods section on page 19.

Figure 4: The task schema does not show a confidence rating for base rates.

Thank you for catching this. We have now added the confidence ratings for base rates to the task in Figure 4b in the revised version of the manuscript. We have furthermore corrected a typo in Figure 4a: The sample size for the group 3 tested in Mai 2021 now indicates 31.

The authors report that base rates are uniformly distributed - this is quite different to other instances of the task where base rates are normally distributed (ideally around the midpoint of the scale). Why this deviation in the design?

We used life events and base rates like those used in past studies of belief updating (Garrett and Sharot 2017, Sharot et al. 2011, Garrett et al. 2017, Korn et al. 2017), which were normal to uniformly distributed (W = 0.952, p = 0.088, Shapiro-Wilk test). The base rates ranged between 10% and 70%, with a mean of 40%. Participants rated their estimates between 3% and 77%, which ensured that for most likely (base rate = 70%) and most unlikely events (base rate = 10%) there was the same space (7%) to update beliefs toward the base rates. Moreover, all statistical models included the absolute estimation errors as a control for variance potentially explained by different estimation error magnitude[42,43]. We added this extra base rate information to the methods section’s task description on page 16.

The task is comprised of only negative life events, which arguably this hinders the generalizability of the results. The authors could mention this as a limitation (there has been a significant quantity of debate about this point in relation to this task: see the work from Ulrike Hahn's lab).

We have added a paragraph to the discussion page 13 to provide a rationale for using only adverse events. This paragraph now reads:

“In this study we tested how actual adverse experiences affect the updating of negative future outlooks in healthy participants and in analogy to studies conducted in depressed patients[19,20,24] following the cognitive model of depression[37]. One open question is whether findings were specific to the adverse event framing[38,39,40]. We argue that under normal, non-adverse contexts belief updating should also be optimistically biased for positive life events, as shown by previous research[41,42]. However, how context such as experiencing a challenging or favorable situation influence the updating of beliefs about positive and negative outlooks remains an open question.”

It would be useful to show the parameter recovery for all parameters (not just the learning rates) and the correlation between parameters (both in simulations and in the fitted parameters).

We apologize for being unclear on this part. The models included two free parameters that were the components of the learning rates: The scaling and the asymmetry parameter. We now have added parameter recovery analyses for the scaling and asymmetry components of the learning rates for (1) the Bayesian model of belief updating during the pandemic, and (2) the RL-like model of belief updating outside the pandemic to the supplement (SI section 2b, SI Figure 4).

Reviewer #2:

The authors investigated how experiencing the COVID-19 pandemic affected optimism bias in updating beliefs about the future. They ran a between-subjects design testing for participants on cognitive tasks before, during, and after lifting the sanitary state of emergence during the pandemic. The authors show that optimism bias varied depending on the context in which it was tested. Namely, it disappeared during COVID-19 and re-emerged at the time of lift of sanitary emergency measures. Through advanced computational modeling, they are able to thoroughly characterize the nature of such alternations, pinpointing specific mechanisms underlying the lack of optimistic bias during the pandemic.

Strengths pertain to the comprehensive assessment of the results via computational modeling and from a theoretical point of view to the notion that environmental factors can affect cognition. However, the relatively small sample size for each group is a limitation.

Thank you for this review.

We acknowledge that sample sizes in each group are lower, especially when breaking down the participant sample into four sub-samples tested in the different contexts. To mitigate concerns we checked the power of the observed context by valence interaction on belief updating. To this aim we simulated new belief updates using the parameters from the best fitting optimistic RL-like model of observed belief updating outside the pandemic, and the rational Bayesian model of observed belief updating during the pandemic. At each iteration we performed a linear mixed effects model analysis of the simulated belief updates[44] analogous to equation iv in the main text. The frequency across 1000 iterations with which the LMEs detected a significant interaction of valence by context on simulated belief updating was 75 %. This frequency indicates the power of the valence by context interaction on observed belief updating. In other words, false negatives were 25% likely, which meant type II errors of failing to reject the null hypothesis when the effect was there. We have added these extra analyses to the main manuscript’s results section page 4 and method’s section page 20.

A major impediment interpreting of the findings is the need for additional measures. While the information on for example, risk perception or the need for social interaction was collected from participants during the pandemic, the fact that these could not be included in the analysis hinders the interpretation of findings, which is now generally based on data collected during the pandemic, for example, reporting increased stress. While authors suggest an interpretation in terms of uncertainty of real-life conditions it is currently difficult to know if that factor drove the effect. Many concurrent elements might have accounted for the findings. This limits understanding of the underlying mechanisms related to changes in optimism bias.

We agree with the reviewer on the limitation arising from the lack of physiological and self-report measures of stress, threat, and perceived uncertainty. To address this point and a similar point raised by reviewer 1 we have added a section to the supplement (SI section 4) that now reports explorative correlations between questionnaire responses of subjective perceptions of risk and anxiety, behavior (e.g. mask wearing, social distancing) and belief updating measured during the 1st strict lockdown.

We now also further discuss this limitation on page 12 of the main text’s discussion.

I recommend that the authors spend more time on explaining the belief-updating task in the presentation of the experiment.

Thank you for this advice. We now provide a clearer and more detailed description of the belief-updating task in the main manuscript’s methods section and have updated Figure 4b to display the confidence rating event in the task schema.

The task description now reads:

“As illustrated in Figure 4b, each of the 40 trials began with presenting an adverse life event. Participants estimated their own risk and the risk of someone else their age and gender. Then the base rate of the event occurring in the general population was displayed on the computer screen. Participants rated their confidence in the accuracy of the presented base rate. Finally, they re-estimated their risk for experiencing the event now informed by the base rate.”

The experimental task seems to include a self-other dimension, which is completely disregarded in the analysis. It would be interesting to explore whether the effect of diluted optimism bias during the pandemic is specific to information about self vs. Other.

We appreciate the reviewer's observation regarding the self-versus-other dimension in the belief updating task design. As now shown in SI Figure 2 the participants indeed displayed an optimism bias: They estimated that adverse events are more likely to happen to others than to themselves (ß = 3.02, SE = 0.86, t (232) = 3.53, p = 5.09e-04, 95% CI [1.33 – 4.71]; SI Figure 2; SI Table 18). This effect was observed overall participants. The pandemic context had no significant effect (ß = -1.91, SE = 3.00, t (232) = -0.64, p = 0.52, 95% CI [-7.82 – 4.00]; SI Table 18). Moreover, following previous studies of optimistically biased belief updating we tested the effect of estimation errors (EE) calculated on the difference between the estimate for someone else (eBR) and the base rate (BR), following: EE = eBR – BR[4,5,25,26]. When categorizing trials as good news or bad news based on this alternative EE calculation the context-by-EE valence interaction remained significant (SI Table 6).

We conclude from these additional analyses that experiencing the pandemic specifically influenced belief updating but did not affect optimism biases in initial beliefs about the future.

Please provide an English translation of the instructions for the task.

We now provide an English translation of the task instructions in the Supplement section 5.

Version 2

Abstract

Optimistically biased belief updating is essential for mental health and resilience in adversity. Here, we asked how experiencing the COVID-19 pandemic affected optimism biases in updating beliefs about the future. One hundred and twenty-three participants estimated the risks of experiencing adverse future life events in the face of belief-disconfirming evidence either outside the pandemic (n=58) or during the pandemic (n=65). While belief updating was optimistically biased and Reinforcement-learning-like outside the pandemic, the bias faded, and belief updating became more rational Bayesian-like during the pandemic. This malleability of anticipating the future during the COVID-19 pandemic was further underpinned by a lower integration of positive belief-disconfirming information, fewer but stronger negative estimations, and more confidence in base rates. The findings offer a window into the putative cognitive mechanisms of belief updating during the COVID-19 pandemic, driven more by quantifying the uncertainty of the future than by the motivational salience of optimistic outlooks.

Introduction

Anticipating the future is an essential part of human thinking. These beliefs guide how we understand the world; updating them is vital for learning to make better predictions and to generalize across contexts. Interestingly, belief updating tends to be optimistically biased1. Even when confronted with negative information, humans often downplay its importance to maintain an optimistic view of the future. The underlying mechanism of optimistically biased belief updating involves an asymmetry in learning from positive and negative belief-disconfirming information2–4, which can unfold in two ways following Reinforcement learning (RL) or Bayes rule5.

Conceptually, Reinforcement learning (RL) and Bayesian models of belief updating are complementary but make different assumptions about the hidden process humans may use to adjust their beliefs when faced with information that contradicts them. The RL models assume that belief updating is proportional to the estimation error. The key idea of the estimation error expresses the difference between how much someone believes for example they will experience a future life event and the actual prevalence of the event in the general population. This difference can be positive or negative. A scaling and an asymmetry parameter quantify the propensity to consider the estimation error magnitude and its valence, respectively. These two free parameters form the learning rate, which indicates how fast and biased participants update their beliefs.

In contrast, Bayesian models assume that following Bayes’ rule the posterior, updated belief is a new hypothesis, formed by pondering prior knowledge with new evidence. The prior knowledge consists in information about the prevalence of life events in the general population. The new evidence comprises various alternative hypotheses. It examines how likely a specific event is to occur or not occur for oneself, compared to the likelihood that it will happen or not happen to others. This probabilistic adjustment of beliefs about future life events can be considered as an approximation of a participant’s confidence in the future. The two free parameters of the Bayesian belief updating model scale how much the initial belief deviates from the updated, posterior belief (i.e., scaling parameter) and the propensity to consider the valence of this deviance (i.e., asymmetry parameter).

Although RL-like and Bayesian updating models make different assumptions about the updating strategy, they are complementary and powerful formalizations of human reasoning. Both models provide insight into hidden, latent variables of the updating process. Most notably, the learning rate and its components, the scaling and asymmetry parameters can vary between individuals and contexts and, through this variance, offer possible explanations for the idiosyncrasy in belief-updating behavior and its cognitive biases.

The COVID-19 pandemic represented a chronic, adverse life context, drastically altering individual and social realities. The existing literature documenting the impact of the COVID-19 pandemic and the associated changes in everyday life on mental health has shown that stress, anxiety, and depression have increased during the pandemic6,7. Moreover, previous work has demonstrated that belief updating during reversal learning became more erratic and was linked to a rise in paranoia during the COVID-19 pandemic across the US8.

However, it is unknown how optimism biases in belief updating about the future and their underlying putative mechanisms changed during the experience of such an unprecedented, adverse life event. Our hypothesis was twofold. We argued that maintaining optimistically biased belief updating under lasting, adverse life conditions is adaptive. The optimism biases are beneficial in exploratory behavior, reduce stress, and improve mental and physical health and well-being9–15. These benefits promote resilience, which is especially important for fitness and survival during a pandemic16,17. On the contrary, optimism biases can lead to suboptimal decision-making. Contextual factors such as acute stress, perceived threat, and depression have been shown to reduce or even reverse optimistically biased belief updating18–24. These findings suggest that optimistically biased belief updating should be weaker when experiencing a pandemic.

We leveraged a belief-updating dataset from 123 participants tested between 2019 and 2022 to rule between these alternative hypotheses. Among them, fifty-eight participants were tested outside the context of the COVID-19 pandemic, either in October 2019, three months before the outbreak in France (n=30) or two years later in June 2022 (n=28), after the lift of the sanitary state of emergency. Their belief updating behavior was compared to sixty-five participants tested during the sanitary state of emergency due to the COVID-19 outbreak in France. This was either during the first very strict lockdown of social and economic life (e.g., schools closed, stay-at-home orders, shops, and museums closed) from March to April 2020 (n=34) or one year later in Mai 2021 (n=31), when lockdown was less strict (e.g., schools open, museums and shops closed, part-time curfew), but the COVID-19 pandemic still unfolding. Belief updating was measured by a behavioral task that asked participants to estimate their risk of experiencing adverse future life events before and after receiving information about these events’ actual base rates. Observed belief updating behavior was fitted to a RL-like and a Bayesian updating model to gain insight into potential underlying strategies of belief updating. The learning rates were compared across groups for insight into how experiencing the COVID-19 pandemic changed beliefs about the future and their updating in the face of belief-disconfirming evidence.

Results

Effects of experiencing the COVID-19 pandemic on optimistically biased belief updating

A linear mixed effects (LME) model was fitted to belief updates to test whether belief updating was less or more biased during the COVID-19 pandemic. The model found a significant interaction estimation error valence by context (β = −5.54, SE = 1.69, t(232) = −3.28, p = 0.001, 95% CI [−8.87 – −2.21]; SI Table 2), which hold when further controlling for the distance variable (SI Table 3). The power of this effect was 75% leaving a 25% risk for type II errors. As shown in Figures 1a and b, optimistically biased belief updating disappeared during the COVID-19 pandemic compared to participants tested outside the pandemic. More specifically, it was decreased among participants tested during the initial COVID-19-related strict lockdown in March and April 2020 (EE valence by context 1: β = −7.39, SE = 2.29, t(228) = −3.32, p = 0.002, 95% CI [−11.91 – −2.86]; SI Table 4), as well as in Mai 2021 (EE valence by context 2: β = −5.59, SE = 2.36, t(228) = −2.37, p = 0.02, 95% CI [−10.24 – −0.93]; SI Table 4), compared to those tested before the outbreak in October 2019, respectively. The bias re-emerged among participants tested one year later at the time of the lift of the sanitary state of emergency in June 2022, returning to levels akin to those observed before the pandemic in October 2019 (EE valence by context 3: β = −2.11, SE = 2.46, t(228) = −0.86, p = 0.39, 95% CI [−6.95 – 2.73]; Figure 1a, SI Table 4). The effect of the COVID-19 pandemic on belief updating was driven by a significant decrease in belief updating following good news during the pandemic compared to participants tested outside the pandemic (t(121) = 2.66, p = 0.009, Cohen’s d = 0.48, two sampled, two-tailed t-test, Figure 1b). No contextual group difference was observed for belief updating following bad news (t(121) = −1.77, p = 0.08, Cohen’s d = −0.32, two sampled, two-tailed t-test, Figure 1b). This effect could be reproduced when fitting an analogous LME to belief updates observed in the group of participants (n=28) who were tested both before and during the pandemic (EE valence by context interaction: β = −7.66, SE = 1.49, t(103) = −5.13, p < 0.001, 95% CI [−10.62 – 4.70]; SI Figure 1, SI Table 5). Moreover, previous studies of optimistically biased belief updating calculated the estimation error (EE) on the difference between the estimate for someone else (eBR) and the base rate (BR), following: EE = eBR − BR4,5,25,26. When categorizing trials as good or bad news based on this alternative EE calculation the context-by-EE valence interaction remained significant (SI Table 6). Note that all effects were controlled for participants age, years of higher education, gender, confidence in the base rates, belief updating task design, and estimation error magnitude.

Behavioral results.

(a) Boxplots display the belief-updating bias (i.e., the difference between the belief update for good news and belief update for bad news) in each of the four participant groups, tested before the pandemic in October 2019 (n=30), during the first lockdown from March to April 2020 (n=34), with less restrictive measures in Mai 2021 (n=31), and at the end of the pandemic in June 2022 (n=28). (b) Belief updating for good and bad news during (n=65) and outside the pandemic (n=58). (c) Confidence ratings, and (d) estimation errors for bad and good news during and outside the pandemic. Boxplots in all panels display 95% confidence intervals, with boxes indicating the interquartile range from Q1 25th to Q3 75th percentile. The horizontal black lines indicate medians and whiskers range from minimum to maximum values and span 1.5 times the interquartile range. The dots correspond to individual participants. The squares in the boxplots in (b) correspond to mean observed updates (purple) and mean modelled updates (blue; averaged across 1000 estimations) from the best-fitting models in each context, which were the optimistically biased RL-like model of belief updating outside and the rational Bayesian model of belief updating during the Covid-19 pandemic. The source data file provides exact p-values. *p < 0.05 two-sampled, two-tailed t-tests, * p < 0.05 two-sampled, one-tailed t-tests.

Effects of experiencing the COVID-19 pandemic on belief updating variables

As shown in Figure 1c, experiencing the COVID-19 pandemic influenced participants’ confidence in the base rates, with significantly lower confidence ratings observed among those tested outside the pandemic compared to those tested during it (β = 14.11, SE = 4.52, t(233) = 3.12, p = 0.002, 95% CI [5.19 – 23.02]; SI Table 7). Moreover, a significant interaction of EE valence by context (β = 2.19, SE = 0.67, t(233) = 3.28, p = 0.001, 95% CI [0.88 – 3.51]; SI Table 8) was found for absolute estimation error magnitude. This finding indicated that participants tested during the pandemic overestimated their risk of experiencing adverse future life events relative to base rates more largely than participants tested outside the pandemic (t(121) = −3.01, p = 0.003, Cohen’s d = −0.54, Figure 1d). On the contrary, the two groups did not differ significantly in the magnitude of negative estimation errors (i.e., initial underestimations relative to base rates; t(121) = −0.49, p = 0.63, Cohen’s d = −0.09, two-sampled, two-tailed t-tests; Figure 1d). This finding contrasts with the observed difference in how often they made positive estimation errors (i.e., the number of good news trials). Participants tested during the pandemic overestimated less frequently than participants tested outside the pandemic (t(121) = 2.40, p = 0.02, Cohen’s d = 0.43, two-sampled, two-tailed t-test). No significant difference between groups was found for the frequency of underestimations (i.e., reflected by the number of bad news trials; t(121) = −1.85, p = 0.07, Cohen’s d = −0.33, two-sampled, two-tailed t-test). These results indicated that participants held fewer but stronger negative future outlooks during the pandemic compared to those tested outside the pandemic.

Computational model comparisons.

Twelve alternative models from RL-like (blue) and Bayesian (orange) updating model families were fitted to observed belief updates for participants tested during the COVID-19 pandemic (left panel columns) and outside the pandemic (right column panels). (a) Protected exceedance probabilities for each of the 12 alternative models, which is the probability that the model predominates in the population above and beyond chance. (b) Posterior model attributions. Colored cells display the probability that individual participants (y-axis) will be best explained by a model version (x-axis). (c) Estimated model frequencies correspond to how many participants are expected to be best described by a model version, with error bars corresponding to standard deviations. The red line indicates the null hypothesis that all model versions are equally likely in the cohort (chance level). Labels on the x-axis of the histogram and bar graphs indicate the model versions with non-silenced parameters (S – scaling, A – Asymmetry) and PR – personal relevance of events.

Effects of experiencing the COVID-19 pandemic on putative mechanisms of belief updating

We then sought to identify which putative strategy participants used to update their beliefs about the future during and outside the pandemic. To answer this question, we used computational modeling and model comparisons to rule between twelve alternative models. This approach revealed that belief updating outside the pandemic was more RL-like and optimistic (pxp = 1, Ef = 0.77), while during the pandemic, it was best explained by a rational Bayesian updating model (pxp = 0.90, Ef = 0.43; Figure 2a-c). Similar findings were obtained when conducting model comparisons in the participants tested both before and during the lockdown (n=28; SI Figure 3).

Effects of experiencing the COVID-19 pandemic on hidden, latent variables of belief updating

Next, we compared the effects of experiencing the COVID-19 pandemic on the learning rates and its components. To show that this adverse context effect was indeed mediated by alterations in asymmetrical learning we compared the scaling and asymmetry parameters obtained from the overall best-fitting model across the whole dataset of n=123 participants. This was Model 1 – the optimistically biased RL-like model of belief updating (Ef = 0.40, pxp =0.99, SI Figure 6).

A linear mixed effects model (LME), analogous to the LME fitted to observed belief updates, was fitted to the learning rates and detected a main effect of EE valence (β = 0.09, SE = 0.01, t(236) = 7.14, p = 1.18e-11, 95% CI [0.06 – 0.11]; SI Table 9), and a significant interaction EE valence by context (β = −0.03, SE = 0.02, t(236) = −2.11, p = 0.04, 95% CI [−0.07 – −0.002]; Figure 3a, SI Table 9). A main effect of EE valence (β = 0.08, SE = 0.02, t(105) = 3.22, p = 0.002, 95% CI [0.03 – 0.12]; SI Table 10) and context (β = −0.10, SE = 0.03, t(105) = −3.10, p = 0.002, 95% CI [−0.17 – −0.04]; SI Table 10) on learning rates was detected when comparing the participants, who were tested both before and during the pandemic. As shown in Figure 3a, all participants’ learning rates were lower in response to bad news than to good news. Still, the difference between good and bad news learning rates was significantly reduced for participants tested during the pandemic. In line with the observed belief updating after good and bad news, the effect of context on the learning rates was driven by a decrease in the learning rates from positive estimation errors in participants tested during the pandemic compared to participants tested outside the pandemic (t(121) = 2.17, p = 0.03, Cohen’s d = 0.39, two-sampled, two-tailed t-test). Both groups did not differ in their learning rates from negative estimation errors (t(121) = 0.87, p = 0.39, Cohen’s d = 0.16, two sampled, two-tailed t-test).

Parameter recovery was successful for the scaling (r = 0.92, p < .001) and asymmetry (r = 0.82, p < .001) components of the learning rates (Figure 3b), which indicated that the model gave identifiable values for these parameters (parameter recovery was also conducted on each group and each model family separately, results are reported in SI section 2b and SI Figure 4). We, therefore, were able to explore potential group differences in the learning rate components in more detail. Linear mixed effects modeling found a main effect of context for the asymmetry component (β = −0.04, SE = 0.02, t(117) = −2.32, p = 0.02, 95% CI [−0.07 – −0.01]; Figure 3c, SI Table 11), but not for the scaling component (β = −0.07, SE = 0.05, t(117) = −1.54, p = 0.13, 95% CI [−0.16 – 0.02]; Figure 3c, SI Table 12). The average asymmetry of learning rates was positive in both groups but significantly smaller in participants tested during the pandemic than those tested outside (t(121) = 2.00, p = 0.048, Cohen’s d = 0.36, two-sampled, two-tailed t-test, Figure 3c). This result indicated that participants considered positive estimation errors more than negative ones but less when experiencing the COVID-19 pandemic. Similar results were found in the within subject group (n = 28), with a significant main effect of context on asymmetry (β = −0.06, SE = 0.02, t(51) = −3.72, p = 0.001, 95% CI [−0.09 – −0.03]; SI Table 13), but not on scaling (β = −0.10, SE = 0.05, t(51) = −1.96, p = 0.06, 95% CI [−0.21 – 0.003]; SI Table 14).

Parameter comparisons between participants tested during (n=65) and outside (n=58) the COVID-19 pandemic.

(a) Learning rates. Boxplots display 95% confidence intervals for learning rates from the RL-like updating model that assumed updating is proportional to the estimation error with an asymmetry and a scaling learning rate component. (b) Parameter recovery for learning rate components of the overall best fitting Model 1 (n=123). Pearson’s correlation between generating and recovered parameters for scaling (left panel) and asymmetry (right panel) learning rate component. r –Pearson’s correlation coefficient against zero. Source data and exact p-values are provided as a Source Data file. (c) Group comparisons for scaling and asymmetry components. Boxplots display 95% confidence intervals for the learning rate’s scaling (left panel) and the asymmetry (right panel) component. Boxes in all boxplots correspond to the interquartile range from Q1 25th percentile to Q3 75th percentile. The horizontal black lines indicate medians and whiskers range from minimum to maximum values and span 1.5 times the interquartile range. The dots correspond to individual participants. *p < 0.05. P-values were obtained with two-sampled, two-tailed t-tests between groups, and exact p-values are provided in the source data file.

Discussion

This study investigated how experiencing the COVID-19 pandemic impacted the optimism biases in updating beliefs about the future. Belief updating was optimistically biased before the COVID-19 outbreak, it faded during the COVID-19 pandemic and reemerged after the pandemic. The lack of optimistically biased belief updating during the pandemic was related to three effects: (1) a decreased sensitivity to positive belief-disconfirming information, (2) fewer but stronger negative beliefs about the future, and (3) more confidence in base rates. Computational modeling showed that belief updating during the pandemic was best described by a rational Bayesian model. In contrast, an optimistic RL-like model best-approximated belief updating outside the pandemic. Both models showed that the attenuated optimistically biased belief updating during the pandemic was not due to a learning deficit. The groups were similar in how much they integrated overall evidence in favor or against initial beliefs. On the contrary, it was explained by a diminished learning asymmetry in considering positive belief-disconfirming evidence that paralleled the observed belief-updating behavior.

The finding that optimistically biased belief updating faded during the pandemic favors the hypothesis that experiencing an adverse life event such as a pandemic weakens optimistic outlooks. It further aligns with the body of research that has explored the malleability of the optimistically biased belief updating and information integration under acute threat and stress and mood disorders such as depression18–24,27. While our findings align with these previous findings, we also observed a difference. Notably, our sample tested during the pandemic considered positive, favorable information less while showing no change in negative, unfavorable information consideration. Differences in populations and task designs might explain these odds. However, it could also be specific to experiencing the COVID-19 pandemic, which involved an immediate, unpredictable, and global health threat with high uncertainty about its outcome and significant psychological repercussions6,28.

Mental health assessments during the COVID-19 pandemic indicated that anxiety, stress, paranoia, and depression levels were more prevalent in the population7,8. The rapid spread of SARS-CoV-2 and the emergence of COVID-19 cases worldwide constitute a challenging situation that, in five months, has shifted from an elusive and distant threat to an immediate and drastic health and economic crisis. All citizens were confronted daily with alarming figures such as infection rates or mortality, and rather mundane everyday activities, from grocery shopping to jogging, became stressful and threatening situations during which one could catch a potentially fatal infection. Moreover, for many, the COVID-19-related lockdowns of social and economic life implied a physical cut-off from friends and relatives, and life plans, routines, and activities were severely disrupted. It also implied a substantial economic risk. Previous research has shown that economic uncertainties, particularly during marked economic inequality and epidemics, can contribute to belief-updating fallacies reflected by the rise of conspiracy theories29.

We did not collect physiological measures of stress or information about the COVID-19 infection status of participants, which precludes a direct exploration of the immediate effects of experiencing the infection on belief-updating behavior and the potential interaction with anxiety and stress levels. Although subjective ratings of the perceived risk of death from COVID-19 correlated negatively to the beliefs updating bias measured during the pandemic, this result was obtained in a subset of participants, retrospectively (SI section 4). We thus cannot directly attribute the observed lack of optimistically biased belief updating during the lockdown to psychological causes such as heightened anxiety and stress. This limitation is noteworthy, as the impact of experiencing the pandemic on belief updating about the future could differ between those who directly experienced infection and those who remained uninfected. It is also important to acknowledge that our study was timely and geographically limited to the context of the COVID-19 outbreak in France. Cultural variations and differences in governmental responses to contain the spread of SARS-CoV-2 may have impacted the optimism biases in belief updating differently.

The observed lack of optimistically biased belief updating may be interpreted as an adaptive response to the experience of an unprecedented level of uncertainty and chronic threat during a global crisis. Although we did not have access to anxiety and stress perceptions during and outside the pandemic, our computational modeling results corroborate to some extent this interpretation. Notably, belief updating was more optimistically biased RL-like outside the pandemic and more rational Bayesian-like during the pandemic. The biased RL-like updating behavior observed outside the pandemic indicated that participants relied on the motivational salience of positive estimation errors to teach them how to update their beliefs about the future by trial and error. This finding aligns with past work, showing that RL-like updating models best explain belief updating in non-threatening, non-stressful, and predictable laboratory contexts5. It further suggests that RL strategies are a computationally efficient way to guide decision-making and belief formation when the environment is stable and predictable30. For instance, in environments with well-defined reward structures, the human brain has been shown to efficiently rely on RL and avoid a computational overhead associated with the Bayesian-like inference process31. On the contrary, belief updating was more rational Bayesian-like during the COVID-19 pandemic, indicating that participants weighed the uncertainty of evidence in favor and against their prior beliefs. This finding aligns with research about Bayesian networks to model semantic knowledge processing under uncertainty32 and with work that uses Bayes rule to understand how humans learn and choose under uncertainty33–35.

It is essential to acknowledge that computational modeling provides insight into potential mechanisms, but this excludes inferences on whether humans indeed update beliefs in the way the best-fitting model assumes. Other models, such as evidence accumulation models, may also work when humans update their beliefs about the future and their immediate surroundings36. Unfortunately, we did not assess reaction times during belief updating, which is crucial for fitting evidence accumulation models such as drift-diffusion models to observed behavior. However, we can infer from our findings that the two model families employed to fit observed belief-updating behavior represented two different but complementary prediction strategies. These strategies were then used to function in the uncertainty of real-life conditions. We call for more studies investigating these computational models’ psychological and biological validity under certainty and uncertainty.

In this study we tested how actual adverse experiences affect the updating of negative future outlooks in healthy participants and in analogy to studies conducted in depressed patients19,20,24 following the cognitive model of depression37. One open question is whether findings were specific to the adverse event framing38–40. We argue that under normal, non-adverse contexts, belief updating should also be optimistically biased for positive life events, as shown by previous research41,42. However, how context such as experiencing a challenging or favorable situation influence the updating of beliefs about positive and negative outlooks remains an open question.

In conclusion, our results provide insight into the resilience and adaptability of belief-updating processes during and following the COVID-19 pandemic. They demonstrate the malleability of the human ability to anticipate the future and how it can adapt to real-life conditions under which an overly optimistic view of future risks would be harmful.

Methods

Ethical considerations

The study protocol followed the Declaration of Helsinki and was approved by the local ethics committee at Sorbonne University. All participants provided informed consent. The authors declare no competing interests.

Participants

One hundred twenty-five participants (mean age = 37.50 ± 1.28, 99 females) allotted to four different groups were recruited for the study (see Table 1 and SI Table 15) via a public advertisement. Two participants from the group tested in June 2022 were excluded from the analyses because they always indicated the same risk estimate for each event.

Experimental design

The first group of 30 participants (mean age = 33.73 ± 1.96, 18 females) was recruited in October 2019 before the COVID-19 outbreak in France (Figure 4a). These participants were tested in the laboratory. A second group of 34 participants (mean age = 42.24 ± 3.34, 25 females) was recruited from March to April 2020 for online testing during the first COVID-19-related lockdown of social and economic life, with schools closed. A third group of 31 participants (mean age = 42.42 ± 3.35, 20 females) was recruited and tested online immediately after the last lockdown and still during the COVID-19 pandemic in Mai 2021. A fourth group of 30 participants (mean age = 34.66 ± 2.71, 16 females) was recruited at the lift of the COVID-19 pandemic-related state of emergency and tested in the laboratory in June 2022 (Figure 4a). This group was also used to rule out an eventual effect of task design. Half of them (n = 15) performed a one-run task design, and the other half (n = 15) performed a two-run task design (e.g., see in more detail the belief updating task description below). Note the 30 participants tested before the COVID-19 pandemic were recontacted during the first strict lockdown to re-perform the belief updating task online (Figure 4a). This allowed us to check for the effects of experiencing a COVID-19-related lockdown within the same cohort of participants. Two of the thirty participants of this group did not respond. Therefore, the sample size for the within-group test-retest analyses was 28 participants.

Experimental design.

(a) Timeline of testing. Four groups were tested, before the COVID-19 outbreak in October 2019, during the first complete lockdown of social and economic life in March and April 2020, after a partial lockdown in Mai 2021, and after the lift of the pandemic-related state of emergency in June 2022. (b) Belief updating task. Panels show subsequent appearances on the screen within a good news trial (left panels) and a bad news trial (right panel). Responses were self-paced. The task goal was to estimate the risk of experiencing different adverse future life events (e.g., tooth decay) for oneself (E1) and for somebody else (eBR) before and after (E2) being presented with information about the event’s prevalence in the general population (i.e., base rate).

Sample sizes:

The sample sizes were determined by a power analysis using the power curve function in R (version 1.2.5033) and building on the good news/bad news bias observed in the first group tested in October 2019 before the COVID-19 outbreak in France. The sample size required to replicate a significant effect of estimation error valence on the updating with a power between 80% and 90% lay between 28 and 35 participants, respectively.

Belief Updating Task

All participants performed a belief-updating task (Figure 4b). For in-person testing, stimulus presentation and response recording were done with the Psychophysics toolbox in MATLAB (R2018b, Update 6, version 9.5.0.1265761). The online testing was done using Qualtrics (Qualtrics Software, version March 2020 of Qualtrics, Copyright 2020 Qualtrics. Available at https://www.qualtrics.com).

The task comprised 40 trials with 40 adverse lifetime events and base rates. In each trial, participants were asked to estimate the likelihood of experiencing an adverse event in the future for themselves and somebody else before and after receiving information about the likelihood of occurrence in the general population (i.e., the base rate). The adverse life events and their actual base rates were taken from previously published work in healthy controls2,43. The base rates for events were normal to uniformly distributed (W = 0.952, p = 0.088, Shapiro-Wilk test). The base rates ranged between 10% and 70%, with a mean of 40%. Participants rated their estimates between 3% and 77%, which ensured that for most likely (base rate = 70%) and most unlikely events (base rate = 10%) there was enough space (7%) to update beliefs toward the base rates42,43. Moreover, all statistical models included the absolute estimation errors as a control for variance potentially explained by different estimation error magnitudes42,43.

In more detail, as illustrated in Figure 4b, each of the 40 trials began with presenting an adverse life event. Participants estimated their own risk and subsequently the risk of someone else their age and gender. Then the base rate of the event for occurring in the general population was displayed on the computer screen. Participants rated their confidence in the accuracy of the presented base rate. Finally, they re-estimated their risk for experiencing the event now informed by the base rate.

The task design varied between some groups:

Fifty-eight participants underwent assessment outside the COVID-19 pandemic, with 45 performing a two-run task design (n=30 tested before the outbreak in October 2019; n = 15 tested at the end of the sanitary state of emergency in June 2022). The remaining 13 participants tested outside the pandemic performed the one-run task design like the 65 participants tested during the pandemic.

In the two-run task design, participants performed a first run of 40 trials. Each trial started with the display of an adverse lifetime event. Participants were asked to estimate the risk of experiencing this event in the future for themselves (E1 rating) and for somebody else (eBR rating). At the end of each trial, they received information about the event’s base rates and rated their confidence. In a second run, they saw an adverse future life event and its base rate on each trial, they then re-estimated their risk (E2 rating) on a trial-by-trial basis.

The one-run task design is displayed in Figure 4b and consists of one run of 40 trials. Within each trial, participants first estimated the risk of experiencing a future adverse lifetime event for themselves (E1 rating) and for somebody else (eBR rating), were presented with the base rate for this event (BR), rated their confidence in the base rate and re-estimated their risk of experiencing the event in the future (E2 rating). Note that all analyses were controlled for these differences in task design, which had non-significant effects on belief updating, confidence ratings, estimation error magnitude, and learning rates (see corresponding SI tables with LME results).

Belief updating task measures of interest:

The estimation error indicated whether participants overestimated or underestimated their likelihood of experiencing an adverse event (E1) relative to its actual base rate (BR). The estimation error (EE) was calculated according to the equation i: EE=E1−BR i.

The estimation error was further used to categorize trials into good or bad news trials:

For good news trials, the estimation error was positive (EE > 0), which indicated an overestimation of one’s likelihood of experiencing an adverse life event relative to the base rate of that event (E1 > BR). For bad news trials, the estimation error was negative (EE < 0), which indicated an underestimation of one’s likelihood of experiencing an adverse event relative to its actual base rate (E1 < BR).

The main variable of interest was the magnitude of belief updating (UPD), which was calculated as the difference between the first (E1) and the second (E2) estimate after receiving information about the base rate (BR). Notably, the update was calculated for good and bad news trials, respectively, following equation ii: if EE>0 UPDgood news=E1−E2if EE<0 UPDbad news=E2−E1 ii.

Lastly, the difference between updating after good and bad news was calculated to assess the updating bias following equation iii: UDB=UPDgood news−UPDbad news iii.

A positive difference indicated that participants updated their beliefs about their lifetime risks of experiencing adverse life events more frequently following good news than bad news.

For each participant, trials that did not receive a response (on average 0.44 trials per subject) and trials with an EE = 0 (on average 0.63 trials per subject) were excluded from the analyses.

The distance measured the extent to which participants consider their probability of experiencing a given adverse event (E1) different from the lifetime risk of someone from a similar socio-economic background (eBR). If positive, it reflected an optimistic bias in initial estimates. The following, distance = eBR – E1, was calculated. Additional analysis to control for this measure was added to the SI (SI Table 3).

Model-free statistical analyses of observed belief updating behavior:

The main aim of this study was to assess how belief updating was affected by the context of experiencing the COVID-19 pandemic.

We, therefore, conducted between-context analyses, contrasting groups tested during (i.e., during the first lockdown in March/April 2020 and immediately after the last lockdown one year later in Mai 2021) and outside the pandemic context (i.e., before the outbreak in October 2019 and one year after the last lockdown in June 2022). All statistical tests were conducted using the MATLAB Statistical Toolbox (MATLAB 2018b, MathWorks) and JASP (JASP 0.16.4).

A first linear mixed effects model (LME 1) was fitted to the belief updating, following equation iv: UPD=β0Intercept+βEE|EE|+βEE valenceEE valence+βcontextcontext+βdesigndesign+βageage+βgender+βeducationeducation+βEE valence*contextEE valence by context+(1|subject)+(1+|EE|subject)+(1+EE valence|subject) iv.

The model included fixed effects for estimation error magnitude (|EE|), estimation error valence (EE valence, coded −1 for bad news trials and 1 for good news trials), context (coded 0 for outside and 1 for during the COVID-19 pandemic), task design (coded 1 for one-run, 2 for two-run design), age, gender (coded 0 for male, 1 for female participants), level of education, and the interaction of interest EE valence by context. The model also included random intercepts nested by subject number and random slopes for estimation error magnitude and valence.

Subsequently, the same Linear Mixed Effects (LME) model was applied again to the belief update to explore the categorical effect of context in conjunction with EE valence. This allowed for a more specific comparison of the impact of EE valence between contexts (groups): those tested before the COVID-19 pandemic outbreak in October 2019 (baseline) compared to those tested during the initial COVID-19-related lockdown (context 1), those tested immediately after the last lockdown during the pandemic in Mai 2021 (context 2), and those tested one year post-pandemic in June 2022 (context 3), respectively.

Post-hoc two-tailed and one-tailed t-tests were conducted to characterize the directionality of detected main effects and interactions.

Posthoc power analysis

The best fitting computational models of belief updating in each context (i.e., during and outside the pandemic) and their free parameters were used to simulate new belief updates44. Simulations were repeated 1000 times. At each iteration, the above-described linear mixed effects model (equation iv) was fitted to the simulated belief updates. The frequency across 1000 iterations with which the LME detected a significant interaction of valence by context on simulated belief updating indicates the power of this interaction effect and the chance for type II errors of failing to reject the null hypothesis when the effect was there.

Model-based analyses of belief-updating behavior

To gain more insight into putative cognitive mechanisms of belief updating during and outside the COVID-19 pandemic, two families of non-linear computational models were fitted to observed belief updating behavior, which is specified below.

Model specifications

(1) Reinforcement learning model of belief updating

A Reinforcement learning-like model assumed that belief updating is proportional to the magnitude of the estimation error following Kuzmanovic and Rigoux, 20175. The learning rate scaled the effect of the estimation error on belief updating following the generic equation v: UPD=LR*EE v.

Importantly, the learning rate was estimated for good and bad news trials separately and following equations vi and vii: LRgood news=S+A vi. LRbad news=S−A vii.

For both types of trials, the learning rate was composed of two components that varied across participants. The scaling parameter (S) measured the extent to which a participant took the prediction error into account when updating beliefs. The asymmetry parameter (A) indicated to what extent the belief updating differed for positive and negative estimation errors. The priors for scaling and asymmetry were untransformed and unbound. The mean of the prior distribution for scaling was set to one. Thus, a scaling of one meant that the updating magnitude equaled the estimation error magnitude. The mean of the prior distribution for the asymmetry parameter was set to zero. An asymmetry parameter value larger than zero meant positively biased updating, whereas an asymmetry parameter smaller than zero meant negatively biased belief updating.

A version of the RL-like model of belief updating took the personal relevance (PR) of presented adverse future life events into account following equations viii and ix: UPDgood news=(S+A)*EE*(1−PR) viii. UPDbad news=(S−A)*EE*(1−PR) ix.

The PR weighed the estimation error (EE) and corresponded to the difference between the estimated base rate (eBR – the estimated risk for somebody else) and the initial estimate (E1 – the estimated risk for oneself). Based on the sign of this difference between eBR and E1, the PR was calculated following equations x to xii: If eBR<E1 PR=eBR−E1eBR−1 x. If eBR>E1 PR=E1−eBR99−eBR xi. If eBR>E1 PR=0 xii.

(2) Bayesian belief updating model

A second family of computational models was fitted to belief updating behavior and assumed that belief updating was proportional or equal to the Bayes rule, following equations xiii and xiv5: UPDgood news=(E1−E2b)*(S+A) xiii. UPDbad news=(E2b−E1)*(S−A) xiv.

The scaling parameter (S) corresponded to the tendency of participants to update their beliefs in response to the presented base rate following Bayes’ rule. A scaling smaller than one (S < 1) indicated lesser belief updating than what was predicted by the Bayes rule, and a scaling larger than one (S > 1) indicated more updating than predicted by the Bayes rule.

The Bayes rule was used to define a Bayesian second estimate (E2b, the updated belief), which was calculated following equations xv and xvi: E2b=Prior*LHR xv. E2b=Prior*LHR1+(Prior*LHR) xvi.

With the Prior = P(BR), corresponding to the base rate (BR) of each event following equation xvii: Prior=BR1−BR xvii.

The Likelihood Ratio (LHR) indicates the probability of the initial estimate (E1) relative to the likelihood of the alternative estimated base rate (eBR) following equation xviii: LHR=E11−BReBR1−eBR xviii.

Alternative models of these two model families (RL and Bayesian) were fitted to the observed belief-updating behavior. Each model alternative represented a different combination of free parameters composing the learning rate to test a total of 12 assumptions about the cognitive process underlying belief updating:

RL model 1. Belief updating is asymmetrical and proportional to the estimation error: S + A varied across participants.

RL model 2. Belief updating is non-asymmetrical and proportional to the estimation error: S varied, A was silent (fixed to zero).

RL model 3. Asymmetrical updating is equal to the estimation error: S is fixed (to one), and A varied.

RL model 4. Updating equals the estimation error: S and A were fixed.

RL model 5. Belief updating is asymmetrical, proportional to the estimation error, and moderated by the personal relevance of events (PR): S + A varied. PR was weighting the EE following equations x., xi. and xii.

RL model 6. Belief updating is non-asymmetrical and proportional to the estimation error moderated by PR: S varied, A was fixed, and PR weighted the EE following equations x., xi. and xii.

RL model 7. Asymmetrical updating equals to the estimation error moderated by PR: S was fixed, A varied, and PR weighted the EE following equations x., xi. and xii.

RL model 8. Updating equals the estimation error moderated by PR: S and A were fixed, and PR weighted the EE following equations x., xi. and xii.

Bayesian model 1. Belief updating is asymmetrical and proportional to Bayes rule: S and A varied.

Bayesian model 2. Belief updating is proportional to a rational Bayes rule: S varied, and A was fixed.

Bayesian model 3. Belief updating equals an asymmetrical Bayes rule: S was fixed, and A varied.

Bayesian model 4. Belief updating equals a rational Bayes rule: S and A were fixed.

Model estimation

Models were estimated following the procedure reported by Kuzmanovic and Rigoux, 20175 and Bottemanne et al., 202224. In short, models were not hierarchical, and parameter estimation was thus less sensitive to differences in group sample sizes. For each participant, optimal scaling and asymmetry parameter values were obtained using Bayesian variational inferences implemented in the VBA toolbox45.

Model comparisons

The free energy approximations for a model’s evidence in each participant were entered into a random effect Bayesian model comparison that yielded the two criteria considered for model selection: the estimated model frequency (Ef) in each group (estimate the frequency of each model in the population) and the protected exceedance probability (pxp), which corresponded to the probability that the hypothesis predominates in the population, above and beyond chance.

Parameter recovery

Parameter recovery analysis was conducted to check whether the free parameters of the winning models were identifiable and described the data better than any other set of parameters. The procedure was the same as reported in Bottemanne et al., 202224. In short, to validate the accuracy of the fitting procedure in providing meaningful parameter values, simulated belief updating data was generated using the observed parameter values for both the optimistic RL model and the optimistic Bayesian updating model. Subsequently, we applied the fitting procedure to these simulated data to iteratively ‘recover’ the parameters. Thereby, the means of the parameters were set to correspond to the observed sample means (i.e., scaling = 0.39 ± 0.02, asymmetry = 0.07 ± 0.01 for the RL model; scaling = 0.42 ± 0.03, asymmetry = 0.05 ± 0.01 for the Bayesian model). This process was iterated to simulate 40 values of belief updates 123 times. The model was then inverted by fitting it to the simulated data, yielding a new set of recovered values for scaling and asymmetry. Finally, the recovered and estimated parameters were compared by assessing their correlation using Pearson’s correlation coefficients.

Parameter comparisons

To compare learning rates and learning rate components across groups, we used the parameters from the optimistically biased RL-like model (RL model 1), which performed best when fitted to the whole dataset (Ef = 0.40, pxp = 0.99) and reproduced the observed updating behavior as shown in SI Figure 6.

Individual learning rates from this RL model 1 and their scaling and asymmetry components were the dependent variables (DV) of the following generic linear mixed effects model (equation xix and xx): LR=β0intercept+βvalencevalence+βcontextcontext+βdesigndesign+βageage+βgendergender+βeducationeducation+βvalence*contextvalence by context+(1+|subject) xix. parameters=β0Intercept+βcontextcontext+βageage+βgendergender+βeducationeducation+(1|subject) xx.

The model included fixed effects for news valence (valence, coded 1 for good news, −1 for bad news), context (coded 0 for outside the pandemic, 1 for during the pandemic), task design (coded 1 for one-run and 2 for two-run), age, gender (coded 0 for male, 1 for female participants), and level of education. It also tested the interaction of interest context by valence. The intercept was nested at random by subject number.

Post-hoc one-sampled and two-sampled t-tests were conducted to characterize the directionality of effects.

Model recovery

To check if the 12 models were identifiable a model recovery analysis was conducted using the VBA toolbox. In more detail, behavior was simulated using each of the 12 models with parameters estimated from participants’ actual behavior. These simulated datasets were then refitted to all alternative models. The model comparison procedure was then performed to evaluate whether each model could accurately recover the parameters that generated the data. This resulted in a 12x12 confusion matrix that compared the performance of all models in fitting each simulated dataset (SI Figure 5). The matrix shows the estimated frequency when fitting to the 12 models (y axis) the behavior generated by each model (x axis) and provided evidence for strong recovery of nearly all models and importantly, the two winning models: the optimistically biased RL-like model and the rational Bayesian model of belief updating. This analysis thus rules out that the two model families were confused and mitigate concerns about the validity of the model selection.